M.Sc. Proposal

### 2010-09-22 notes/proposal.tex
Proposal

\title{M.Sc Proposal}
\author{Roman Zeyde}
\begin{document}
\maketitle
\section{Introduction}
Electrokinetics describes dynamics of particles in ion-containing fluids.
When a particle acquires surface charge, a thin layer of ions in the fluid of opposite charge is attracted to the surface via electric forces, creating a double-layer structure (see Figure \ref{fig:EDL}).
This structure, called ``Debye layer'', electrically screens the surface charge, thus creating a potential difference between the particle and the outer layer of
the fluid bulk, called ``zeta potential''.

\begin{figure}[htbp]
\begin{framed}
    \begin{center}
        \includegraphics[width=0.25\textwidth]
            {ElectricDoubleLayer.eps}
        \caption{Electrical Double Layer is composed of (1) particle's surface, (2) Debye layer and (3) fluid bulk.
        Zeta potential is defined as the voltage between (1) and (3).}
        \label{fig:EDL}
    \end{center}
\end{framed}
\end{figure}

In case the layer width is much smaller than the particle's size an analytical solution in the Debye-layer can be found \cite{Yariv10a}, and be used as the appropriate boundary condition for the macroscale bulk problem on the particle's surface.

Electrokinetics is described by the following physical phenomena (in dimensionless formulation):
\begin{enumerate}
\item Electrostatic potential - the fluid bulk contains no free charge:
\begin{equation} \label{eq:Laplace}
    \bnabla \cdot (C \bnabla \varPhi) = 0.
\end{equation}
\item Fluid dynamics - Stokes incompressible flow with electrostatic force component:
\begin{equation} \label{eq:Stokes}
\left\{ \begin{array}{l}
\bLaplacian \bV - \bnabla P + \Laplacian \varPhi \bnabla \varPhi = 0 \\
\bnabla \cdot \bV = 0 \end{array} \right.
\end{equation}
\item Ions' diffusion and advection (via fluid's velocity field):
\begin{equation} \label{eq:Nernst}
\Laplacian C - \alpha \bV \cdot \bnabla C = 0
\end{equation}
\end{enumerate}

The boundary conditions correspond to the specific problem at hand and are usually coupled as well as the partial differential equations themselves.

The closed form approximate solution has been developed for spherical particle
and small electric field \cite{Yariv10b} but since the equations are coupled
and non-linear, it is hard to extend the analytic solution to more general
settings.

\section{Research goals}
Our goal is to implement an accurate and fast iterative numerical solver for electrokinetic problems, in cases there are no closed-form solutions known (e.g. asymmetric particles or large electric fields).

@@

Moreover, the convergence rate of the solver will be optimized using numerical acceleration methods to achieve high accuracy in reasonable solver's execution time.

Achieved results will provide a numeric model for interesting small-scale physical phenomena, such as electrophoretic\footnote{Electrophoresis is
the physical phenomenon of particles motion in an ion-containing fluid under the influence of an electric field.} autonomous micro-swimmers \cite{Paxton04, Howse07} by numerically modeling an appropriate particle's configuration.

\section{Achieved results}
The numerical solver is being implemented as a MATLAB package that
provides the needed tools for numerical solution of an electrophoresis problem.

The solver applies iterative relaxations to solve the coupled system, by
iterating the equations and relaxing repeatedly each one of them
for the corresponding variable:
\begin{itemize}
  \item Laplace equation \ref{eq:Laplace} is used to update $\varPhi$.
  \item Stokes equation \ref{eq:Stokes} is used to update $\bV$ and $P$.
  \item Nernst-Planck equation \ref{eq:Nernst} is used to update for $C$.
\end{itemize}
While solving specific equation for a specific variable, all other variables keep their values from previous iteration, resembling ``Gauss-Seidel'' method for non-linear systems.

Problem's configuration is taken from \cite{Yariv10b} and consists of a spherical particle, surrounded by infinite Stokes fluid.
Suppose that the particle is stationary and fluid flows around it, slipping across the boundary layer (due to ``Debye layer'' effects) with a velocity field denoted by $\bV$.

Spherical coordinates $(R,\theta,\phi)$ system is chosen to make the problem axisymmetric around the velocity vector of the fluid, therefore independent of $\phi$.
The electrokinetic problem is written as two-dimensional partial differential equation system for $R \in [1,\infty)$ and $\theta \in [0, \pi]$.
Natural grid choices are a logarithmic grid for $R$ and an uniform grid for $\theta$.

The following linear operators are implemented in spherical coordinates on the grid described above: Scalar Laplacian (for $\varPhi$ and $C$), Vector Laplacian (for $\bV$), Gradient (for $\Phi, C$ and $P$) and Divergence (for $\bV$).

Each differential operator is written in spherical coordinates on the regular $(R, \theta)$ grid and being linear -- is represented as a sparse matrix, making the algebraic manipulations much more straightforward.

The boundary conditions are computed for $\theta = 0$ and $\theta = \pi$ using symmetry considerations. At $R = \infty$, we assume constant fluid velocity and $\bV = \cal{U} \bzhat$, constant electric field $\bnabla \varPhi = -\beta \bzhat$ and constant ionic concentration $C = 1$.
The boundary conditions on $R = 1$ are defined by the boundary layer behavior as described in \cite{Yariv10b}, using Dukhin-–Derjaguin slip formula for $\bV$ and ion-selectivity condition for $\varPhi$ and $C$.

After substitution of boundary conditions, each differential equation is transformed to a sparse linear system of the form $\matr{A}\bx = \vect{b}$.
Due to coupling, it is not necessary to find the exact solution -- so we apply few relaxations on each linear system, using an appropriate preconditioner $\matr{M}_\matr{A}^{-1}$:
\begin{eqnarray} \label{eq:Precond}
  \bx_{n+1} &=& \bx_{n} + \matr{M}_\matr{A}^{-1} \left(\vect{b} - \matr{A}\bx_{n}\right)
\end{eqnarray}

After the solver has converged to a solution for the coupled system, the stress tensor $\mathbb{T}$ on the particle $\cl{S} = \left\{\bx : \|\bx\| = 1\right\}$ is computed  (combining Newtonian and Maxwell stresses) and integrated to yield the total force:
\begin{eqnarray}
  \mathbb{T} &=& - P \mathbb{I} + \bnabla \bV + (\bnabla \bV) ^T  + \bnabla \varPhi \bnabla \varPhi - \frac{\bnabla \varPhi \cdot \bnabla \varPhi}{2} \mathbb{I}, \\
  \vect{F} &=& \int_{\cal{S}} \left(\mathbb{T} \cdot \bnhat \right) ds = \boldsymbol{0}.
\end{eqnarray}
Since the problem is axisymmetric, the force should be aligned with the symmetry axis: $\vect{F} = F \bzhat$.
Moreover, since the particle is stress-free in steady flow, we have $F = 0$ for steady-state velocity $\cal{U}$.

This observation enables us to find the correct $\cal{U}$ for given $\beta$, using one-dimensional root-finding algorithm on $F_\beta(\cal{U})$, as a function of $\cal{U}$.

@@

$C$, $\varPhi$ and $P$ are discretized using a central grid whereas staggered grid is used for $\bV$ at the problem domain $\cl{D}$ (see Figure \ref{fig:Grids}).
\begin{figure}[htbp]
\begin{framed}
    \begin{center}
        \includegraphics[width=1\textwidth]
            {StaggeredGrid.eps}
        \caption{Grids for $C$, $\varPhi$, $P$ and $\bV$.}
        \label{fig:Grids}
    \end{center}
\end{framed}
\end{figure}

Finite volume method is used to represent the partial differential equations in the form of algebraic system, using
central differences or upwind scheme (for better stability).

``Ghost points'' are used to define Dirichlet and Neumann boundary conditions on
problem domain's boundary $\partial \cl{D}$.

After boundary counditions are eliminated, Jacobi and Red-Black Gauss-Seidel preconditioners are used for equations \ref{eq:Laplace} and \ref{eq:Nernst}, where Vanka-type \cite{Vanka86} preconditioner is used for equation \ref{eq:Stokes}.
The preconditioners are computed as sparse matrices, making the relaxations (at equation \ref{eq:Precond}) much more straightforward.

Vector extrapolation methods \cite{Sidi90} have been implemented and are used to accelerate solver's convergence.
Multigrid solver \cite{Brandt77, Yavneh06} implementation will be a part of the research.

>>>>

\title{M.Sc. Proposal}
\author{Roman Zeyde}
\begin{document}
\maketitle
\section{Introduction}
Electrokinetic theory describes the dynamics of charged particles in ionic fluids.
When a particle acquires surface charge, a layer of ions of opposite charge is attracted to the surface via electric forces, creating a double-layer structure around the particle (see Figure \ref{fig:EDL}).
This structure, called ``Debye layer'', electrically screens the surface charge, thus creating a potential difference between the particle and the outer layer of the fluid bulk.

\begin{figure}[htbp]
\begin{framed}
    \begin{center}
        \includegraphics[width=0.25\textwidth]
            {ElectricDoubleLayer.eps}
        \caption{Schematic structure of the double layer:
        (1) particle's surface, (2) Debye layer and (3) fluid bulk.
        If the the particle's surface has positive (blue) surface charge, it attracts negative (red) ions from the fluid making the Debye layer be negatively charged (as opposed to the rest of the fluid bulk, which is electrically neutral).
        Zeta potential is defined as the voltage drop on (2).}
        \label{fig:EDL}
    \end{center}
\end{framed}
\end{figure}

In cases where the layer width is much smaller than the particle's size, an analytical solution for the Debye layer dynamics can be found \cite{Yariv10a}.

Electrokinetics is described by the following physical phenomena (in dimensionless formulation):
\begin{enumerate}
\item Electrostatic potential - the fluid bulk contains no free charge:
\begin{equation} \label{eq:Laplace}
    \bnabla \cdot (C \bnabla \varPhi) = 0.
\end{equation}
\item Fluid dynamics - Stokes incompressible flow with an electrostatic force component:
\begin{equation} \label{eq:Stokes}
\left\{ \begin{array}{l}
\bLaplacian \bV - \bnabla P + \Laplacian \varPhi \bnabla \varPhi = 0 \\
\bnabla \cdot \bV = 0 \end{array} \right.
\end{equation}
\item Ion diffusion and advection (via the fluid's velocity field):
\begin{equation} \label{eq:Nernst}
\Laplacian C - \alpha \bV \cdot \bnabla C = 0
\end{equation}
\end{enumerate}
The boundary conditions correspond to the specific problem at hand and defined by particle's geometry, chemical characteristics and fluid's dynamics.

The partial differential equations above are coupled and non-linear. In general, they have no analytic solution. Moreover, any numerical solver must handle the scale disparity, caused by the Debye layer width being much smaller than the particle's size.
A closed-form approximate solution has been developed for spherical ion-exchanging particles and small electric field \cite{Yariv10b} but it is hard to extend the analytic solution to more general systems.

\section{Research goals}
Our goal is to implement an accurate and fast iterative numerical solver for electrokinetic problems, which could be successfully applied for systems that have no closed-form solutions known (e.g. asymmetric particles or large electric fields).

@@

Moreover, the convergence rate of the solver will be optimized using numerical acceleration methods in order to achieve high accuracy in reasonable execution time.

The solver will provide a numerical model for interesting electrokinetical phenomena which have no closed-form analytic solutions yet, such as large field dynamics and electrophoretic\footnote{Electrophoresis is the physical phenomenon of particles motion in an ion-containing fluid under the influence of an electric field.} autonomous micro-swimmers \cite{Paxton04, Howse07}.

\section{Achieved results}
The numerical solver is being implemented as a MATLAB package, which provides the needed tools for numerical solution of an electrophoresis problem.

The solver applies iterative relaxations to solve the coupled system, by
iterating the equations and relaxing each one of them
for the corresponding variable:
\begin{itemize}
  \item Laplace equation \ref{eq:Laplace} is used to update $\varPhi$.
  \item Stokes equation \ref{eq:Stokes} is used to update $\bV$ and $P$.
  \item Nernst-Planck equation \ref{eq:Nernst} is used to update for $C$.
\end{itemize}
While solving a specific equation for a specific variable, all other variables maintain their values from previous iteration, resembling ``Gauss-Seidel'' method for linear systems.

The problem's configuration is taken from \cite{Yariv10b} and consists of a spherical ion-exchanging particle, surrounded by infinite Stokes fluid.
Spherical coordinates $(R,\theta,\phi)$ system is chosen to make the problem axisymmetric, therefore independent of $\phi$.
The electrokinetic problem is written as two-dimensional partial differential equation system for $R \in [1,\infty)$ and $\theta \in [0, \pi]$.
Natural grid choices are a logarithmic grid for $R$ and an uniform grid for $\theta$.

The following linear operators are implemented in spherical coordinates on the grid described above: Scalar Laplacian (for $\varPhi$ and $C$), Vector Laplacian (for $\bV$), Gradient (for $\Phi, C$ and $P$) and Divergence (for $\bV$).
Each differential operator is written in spherical coordinates on the regular $(R, \theta)$ grid and being linear, is represented as a sparse matrix -- simplifying the algebraic manipulations.

The boundary conditions are defined for $\theta = 0$ and $\theta = \pi$ using symmetry considerations. At $R = \infty$, we assume constant fluid velocity $\bV = \cl{U} \bzhat$, constant electric field $\bnabla \varPhi = -\beta \bzhat$ and constant ionic concentration $C = 1$.
The boundary conditions on $R = 1$ are defined by the Debye layer's behavior as described in \cite{Yariv10b}, using the Dukhin-–Derjaguin slip formula for $\bV$ and ion-selectivity for $\varPhi$ and $C$.

After substitution of boundary conditions, each differential equation is transformed into a sparse linear system of the form $\matr{A}\bx = \vect{b}$.
Due to coupling, it is not necessary to find the exact solution -- so we apply few relaxations (\ref{eq:Relax}) on each linear system, using an appropriate preconditioner $\matr{M}_\matr{A}^{-1}$:
\begin{eqnarray} \label{eq:Relax}
  \bx_{n+1} &=& \bx_{n} + \matr{M}_\matr{A}^{-1} \left(\vect{b} - \matr{A}\bx_{n}\right)
\end{eqnarray}

After the solver has converged to a solution for the coupled system, the total stress tensor $\mathbb{T}$ on the particle $\cl{S} = \left\{\bx : \|\bx\| = 1\right\}$ is computed (\ref{eq:Stress}) and integrated to yield the total force (\ref{eq:Force}):
\begin{eqnarray}
  \label{eq:Stress}
  \mathbb{T} &=& - P \mathbb{I} + \bnabla \bV + (\bnabla \bV) ^T  + \bnabla \varPhi \bnabla \varPhi - \frac{\bnabla \varPhi \cdot \bnabla \varPhi}{2} \mathbb{I}, \\
  \label{eq:Force}
  \vect{F} &=& \int_{\cl{S}} \left(\mathbb{T} \cdot \bnhat \right) ds = \boldsymbol{0}.
\end{eqnarray}
Since the problem is axisymmetric, the force should be aligned with the symmetry axis: $\vect{F} = F \bzhat$.
Moreover, since the particle is stress-free in steady flow, we have $F = 0$ for steady-state velocity $\cl{U}$.

This observation enables us to find the correct $\cl{U}$ for given $\beta$, using one-dimensional root-finding algorithm on $F_\beta(\cl{U})$, as a function of $\cl{U}$.

The numerical solution for $\cl{U}$ as a function of $\beta$ is compared to the closed-form solution from \cite{Yariv10b}.
Using a $60 \times 15$ grid for $(R,\theta) \in [1, 100] \times [0, \pi]$, we apply $2\mathord{,}000$ iterations in order to find the total force acting on the particle.
This step is repeated a few times to find the correct $\cl{U}$ so that $F_\beta(\cl{U}) = 0$ for various $\beta$ values -- see Figure \ref{fig:Results} for numerical results.
\begin{figure}[htbp]
\begin{framed}
    \begin{center}
        \includegraphics[width=0.45\textwidth]
            {convergence.eps}
        \includegraphics[width=0.45\textwidth]
            {comparison.eps}
        \caption{Left: Residual $\ell_2$ norm graph. Right: Particle Velocity $\cl{U}$ as a the function of electric field $\beta$.
        The error between the theoretical and the solver's results is less than 2\%.}
        \label{fig:Results}
    \end{center}
\end{framed}
\end{figure}

@@

$C$, $\varPhi$ and $P$ are discretized using a central grid whereas a staggered grid is used for $\bV$ at the problem domain $\cl{D}$ (see Figure \ref{fig:Grids}).
\begin{figure}[htbp]
\begin{framed}
    \begin{center}
        \includegraphics[width=1\textwidth]
            {StaggeredGrid.eps}
        \caption{Grids for $C$, $\varPhi$, $P$ and $\bV$.}
        \label{fig:Grids}
    \end{center}
\end{framed}
\end{figure}

Finite volume method is used to represent the partial differential equations in the form of an algebraic system, using central differences or an upwind scheme (for better stability).

``Ghost points'' are used to define Dirichlet and Neumann boundary conditions on
the problem domain's boundary $\partial \cl{D}$.

After boundary counditions are eliminated, Jacobi and Red-Black Gauss-Seidel preconditioners are used for equations \ref{eq:Laplace} and \ref{eq:Nernst}, where a Vanka-type \cite{Vanka86} preconditioner is used for equation \ref{eq:Stokes}.
The preconditioners are computed as sparse matrices, making the relaxations (at equation \ref{eq:Relax}) much more straightforward.

Vector extrapolation methods \cite{Sidi90} have been implemented and are used to accelerate the solver's convergence.
Multigrid solver \cite{Brandt77, Yavneh06} implementation will also be a part of the research.
